---
title: "Transform Data"
author: "Jerri Schorr"
date: 09-24-2019
output: github_document
---

```{r setup}
library(tidyverse)
#install.packages("babynames")
library(babynames)
babynames # this is a package of JUST data
```

# dplyr

* **d** stands for data
* **plyr** stands for plyers because we manipulate data

## Review

* Try to run the code and see what graph you get

```{r}
ggplot(data = babynames) + 
  geom_line(mapping = aes(x = year, y = prop))
```
    + This graph isn't really helpful because it has too much info 
    + We need to figure out how to extract specific data so that we get a clean visual
    
* In order to use dplyr to the fullest we need to use **tidy data**
    + columns are variables
        + we can select, mutate, and summarize variables
    + rows are observations or cases
        + we can filter, arrange, and group cases

## Your Turn 1

* Select
    + first argument is the data
    + everything after is the subset you would like to pull out

Alter the code to select just the `n` column:

```{r}
select(babynames, name, prop)
#changed code to n only
select(babynames, n)
```

## Quiz

Which of these is NOT a way to select the `name` and `n` columns together?

* option 4 will **not** select name and n

```{r}
#1
select(babynames, -c(year, sex, prop)) # this will select everythin BUT these columns
#2
select(babynames, name:n) # selects a range of columns
#3
select(babynames, starts_with("n")) # this is picking any column that starts with n
#4
select(babynames, ends_with("n")) # this will select columns that end with n
```

## Your Turn 2

* Values go in quotes, Varibles DO NOT

Use `filter`, `babynames`, and the logical operators to find:

* All of the names where prop is greater than or equal to 0.08  
* All of the children named "Sea"  

```{r}
filter(babynames, prop >= 0.07)
filter(babynames, name == "Jerri")

```

* Filter gives you rows that fullfill your arguments
    
## Your Turn 3

Use Boolean operators to return only the rows that contain:

* _Boys_ named Sue  
* Names that were used by exactly 5 or 6 children in 1880  
* Names that are one of Acura, Lexus, or Yugo
    + &, |, xor(a,b), !

```{r}
filter(babynames, sex == "M" & name == "Sue")
filter(babynames, n == 5 | n == 6 & year == 1880 )
filter(babynames, name == "Acura" | name == "Lexus" | name == "Yugo")
# similar way to call all three names
filter(babynames, name %in% c("Acura", "Lexus", "Yugo"))
```

## Help Me

What is the smallest value of n?
What is the largest?

* arrange()
    + first argument is data
    + second argument is the variable column
    + third argument is to settle ties in the data
    + default is small to large
    + desc() makes it go the opposite
    + strings will be arranged alphabetically

```{r}
arrange(babynames, n) #5
arrange(babynames, desc(n)) #99686
arrange(babynames, n, prop) 
```

## Your Turn 4

* Pipe function %>%
    + take output from one line and add it to the next 
    + allows us to do multiple steps in a more organized way

Use `%>%` to write a sequence of functions that: 

1. Filters babynames to just the girls that were born in 2017, *then...*  
2. Selects the `name` and `n` columns, *then...*  
3. Arranges the results so that the most popular names are near the top.

```{r}
#long way 
boys_2015 <- filter(babynames, year == "2015", sex == "M")
boys_2015 <- filter(boys_2015, name, n )
boys_2015 <- arrange(boys_2015, desc(n))
boys_2015

# pipe way
filter(babynames, year == 2015, sex == "M") %>%
  select(name, n) %>%
  arrange(desc(n))
```

## Your Turn 5 - Exam

1. Trim `babynames` to just the rows that contain **your** `name` and **your** `sex`  
2. Trim the result to just the columns that will appear in your graph (not strictly necessary, but useful practice)  
3. Plot the results as a line graph with `year` on the x axis and `prop` on the y axis

```{r}


```

## Your Turn 6

Copmplete the code below to extract the rows where `name == "Khaleesi"`. Then use `summarise()` and `sum()` and `min()` to find:

1. The total number of children named Khaleesi
2. The first year Khaleesi appeared in the data

*(Hint: Be sure to remove each `_` before running the code)*

```{r}
babynames ___ 
  filter(_______________________) ___
  ___________(total = ________, first = _______)
```

## Your Turn 7

Use `group_by()`, `summarise()`, and `arrange()` to display the ten most popular names. Compute popularity as the *total* number of children of a single gender given a name.

*(Hint: Be sure to remove each `_` before running the code)*

```{r}
babynames %>%
  _______(name, sex) %>% 
  _______(total = _____(n)) %>% 
  _______(desc(_____))
```


## Your Turn 8

Fill in the blanks to: 
Rank each name _within its year and sex_. 

*(Hint: Be sure to remove each `_` before running the code)*

```{r}
babynames %>% 
  group_by(____, ____) %>% 
  ______(rank = min_rank(____(prop)))
```


***

# Take aways

* Extract variables with `select()`  
* Extract cases with `filter()`  
* Arrange cases, with `arrange()`  

* Make tables of summaries with `summarise()`  
* Make new variables, with `mutate()`  
* Do groupwise operations with `group_by()`

* Connect operations with `%>%`  


